### 2016-09-17 thesis/intro/intro.tex
tmp commit

% This is taken from the committee report (I think it's a good intro to the intro)
% TODO: past tense...??
It is now commonly understood that cancers are caused by dysregulation of various mechanisms or pathways that allow tumour cells to prolilferate, survive and migrate.
There have been many reports on the association between obesity and worsened cancer prognosis, suggesting that there may be distinct changes that underlie tumorigenesis and/or cancer progression that are specific to patients with higher \gls{bmi}.

% TODO: in the aims section, or here?
In brief, this research aims to determine whether gene expression signatures exist that are specific to obesity across multiple cancer types, and to investigate whether there aare any common pathways being dysregulated in cancers based on these genetic signatures.
Better understanding of the pathways being dysregulated in cancer cells in obese patients may lead to improved clinical decisions, and could thus contribute towards personalised treatment in the future.

@@

\subsubsection{ \textbf{\citet{Creighton2012} study}}
\label{ssub:creighton_study}

\subsubsection{ \textbf{\citet{Fuentes-Mattei2014} study}}
\label{ssub:fuentes_mattei_study}

>>>>

@@

\subsubsection{\textbf{\citet{Creighton2012} study}}
\label{ssub:creighton_study}

\subsubsection{\textbf{\citet{Fuentes-Mattei2014} study}}
\label{ssub:fuentes_mattei_study}

### 2016-09-17 thesis/methods/methods.tex
tmp commit

Samples were classified based on the \gls{who} definition, as shown in Table~\cref{tab:whobmiclass}.
\begin{table}[hb]
	\caption{\gls{who} defined \gls{bmi} classification}
	\label{tab:whobmiclass}
	\begin{center}
		\begin{tabular}{lc}
			\bfseries {Classification} & \bfseries {\gls{bmi} Value}\\
			\hline
			\rule{0pt}{2.25ex}Underweight & \textless{} 20.0\\
			Normal weight/lean & 20.0$\sim$24.9\\
			Overweight & 25.0$\sim$29.9\\
			Obese & \textgreater{} 30.0\\
		\end{tabular}
	\end{center}
\end{table}

@@

The original raw data from the \citet{Creighton2012} and \citet{Fuentes-Mattei2014} studies were downloaded from the \gls{geo} website.
Data related to multiple different cancer types were downloaded from free and publicly available sources such as \gls{tcga} and the \gls{icgc}.

\subsection{Raw data from the \citet{Creighton2012} study}
\label{subsec:crrawdat}

The raw microarray gene expression data files of all *** samples from the \citet{Creighton2012} study were downloaded from the \gls{geo} database (\gls{geo} accession ID: GSE24185).
Clinical data of the samples were obtained from the supplementary table 1 from their paper.
%TODO: add data/sample composition -- ethinicity, grade, etc -- maybe in table form
799 obesity associated gene probes identified in the \citet{Creighton2012} study were obtained from the supplementary data file 1 from their paper.
The raw expression data, clinical data, and the list of gene probes were imported into \gls{r}/RStudio.

\subsection{Raw data from the \citet{Fuentes-Mattei2014} study}
\label{subsec:fmrawdat}

@@

@@

\gls{rnaseq} data for the eight cancer types identified in \cref{subsec:tcga} were downloaded from the \gls{icgc} database.
The \gls{rnaseq} data for each of the eight cancer types were imported into \gls{r}.

@@

\subsection{Creighton data}
\label{subsec:crdatproc}

% \subsubsection{\textbf{Data processing}}

\subsubsection{\textbf{Raw data normalisation}}

The raw microarray data for all 103 samples from \cref{subsec:crrawdat} were normalised using the \gls{rma} method from the \textit{affy} package in \gls{r}.
\gls{rma} normalisation was used to account for the inter-array differences that may have arisen during the data collection process.

\subsubsection{\textbf{Raw data standardisation}}

The data was standardised so that each gene had a mean ($\mu$) of 0 and a standard deviation ($\sigma$) of 1.
This ensured that all of the genes were on the same scale as one another, allowing for comparison across all the genes.

\subsubsection{\textbf{Gene probe ID conversion}}

@@

\subsubsection{\textbf{Sample \gls{bmi} calculation}}

\gls{bmi} of the samples were calculated from the clinical data using \cref{eq:bmicalc}.

\subsection{Fuentes-Mattei data}
\label{subsec:fmdatproc}

% \subsubsection{\textbf{Data processing}}

\subsubsection{\textbf{Raw data normalisation}}

The raw microarray data for all *** samples from \cref{subsec:fmrawdat} were normalised using the \gls{rma} method from the \textit{affy} package in \gls{r}.
\gls{rma} normalisation was used to account for the inter-array differences that may have arisen during the data collection process.

\subsubsection{\textbf{Raw data standardisation}}

The data was standardised so that each gene had a mean ($\mu$) of 0 and a standard deviation ($\sigma$) of 1.
This ensured that all of the genes were on the same scale as one another, allowing for comparison across all the genes.

\subsubsection{\textbf{Gene probe ID conversion}}

Where appropriate, the gene probe IDs in the raw data were converted into their corresponding gene symbols using the \textit{hgu133a.db} package in \gls{r}.
Where there was an overlap in the gene symbols, a single data was chosen to represent that particular gene symbol by using the \texttt{collapseRows} function in the \textit{WGCNA} package in \gls{r}.

\subsection{\gls{tcga}}
\label{subsec:datproctcga}

\subsubsection{\textbf{Raw data normalisation}}

The raw \gls{rnaseq} data for each cancer types from \cref{subsec:tcga} had 1 added, and then logged to the power of 10.

\subsubsection{\textbf{Raw data standardisation}}

The normalised data was standardised so that each gene had a mean ($\mu$) of 0 and a standard deviation ($\sigma$) of 1.
This ensured that all of the genes were on the same scale as one another, allowing for comparison across all the genes.

\subsection{\gls{icgc}}
\label{subsec:datprocicgc}

\subsubsection{\textbf{Raw data format conversion}}
%TODO: add in how I processed the ICGC data format into genes by samples.
The raw \gls{rnaseq} data for each cancer types were formatted so that the rows corresponded to the genes and the columns to the samples.
% TODO: find out what functions I used to reformat the icgc data
This was done using the \textit{dplyr} package in \gls{r}.

\subsubsection{\textbf{Sample ID conversion}}
%TODO: add details on how I stripped the ICGC ID into TCGA ID
%TODO: check the actual ID stripping process
%TODO: check the WGCNA function and check if it uses median/mean/whatever for choosing the sample
\gls{icgc} sample ID in the \gls{rnaseq} data were stripped so that the ID matched with the \gls{tcga} ID in the clinical data.
After the IDs were stripped, samples were checked for any overlaps.
Where there was an overlap in the sample ID in the \gls{rnaseq} data, a single sample was chosen to represent that particular ID by using the \texttt{collapseRows} function in the \textit{WGCNA} package in \gls{r}.

\subsubsection{\textbf{\gls{icgc} sample exclusion criteria}}
\label{subsec:icgcsampleexclusion}

% \subsubsection{\textbf{Sample BMI availability}}

Some of the samples did not have either height or weight data, or both.
These samples were removed from the analyses.
In each of the eight cancer types, the \gls{tcga} sample ID in the clinical data were cross-checked with the \gls{tcga} sample ID in the \gls{rnaseq} data, and vice versa.
Any sample that did not have either clinical or \gls{rnaseq} data were removed.
This ensured that both clinical and \gls{rnaseq} data were available for all of the samples that were included in the analyses.
See \cref{tab:samplesize} for a summary of the number of samples included in the analyses for each cancer type.

@@

@@

\subsection{Box plots}
\label{sub:box_plots}

\subsection{Scatter plots}
\label{sub:scatter_plots}

>>>>

Samples were classified based on the \gls{who} definition, as shown in \cref{tab:whobmiclass}.
\begin{table}[hb]
	\caption{\gls{who} defined \gls{bmi} classification}
	\label{tab:whobmiclass}
	\begin{center}
		\begin{tabular}{lc}
			\bfseries {Classification} & \bfseries {\gls{bmi} Value}\\
			\hline
			\rule{0pt}{2.25ex}Underweight & \textless{} 20.0\\
			Normal weight/lean & 20.0$\sim$24.9\\
			Overweight & 25.0$\sim$29.9\\
			Obese & \textgreater{} 30.0\\
		\end{tabular}
	\end{center}
\end{table}

@@

% TODO: add Cris Print's paper
The raw microarray data from \citet{Creighton2012}, \citet{Fuentes-Mattei2014} and \citet{Gatza2010a}  studies were downloaded from the \gls{geo} website.
\gls{rnaseq} and clinical data for multiple different cancer types were downloaded from free and publicly available sources such as \gls{icgc} and \gls{tcga}, respectively.

\subsection{Raw data from \citet{Creighton2012} study}
\label{subsec:crrawdat}

%TODO: add data/sample composition -- ethinicity, grade, etc -- maybe in table form

The raw microarray gene expression data files of all 103 samples from the \citet{Creighton2012} study were downloaded from the \gls{geo} database (\gls{geo} accession ID: GSE24185).
Clinical data of the samples were obtained from the supplementary table 1 from \citet{Creighton2012} paper.
799 obesity associated gene probes identified in the \citet{Creighton2012} study were obtained from the supplementary data file 1 from \citet{Creighton2012} paper.
The raw expression data, clinical data, and the list of gene probes were imported into \gls{r}/RStudio.

\subsection{Raw data from \citet{Fuentes-Mattei2014} study}
\label{subsec:fmrawdat}

%TODO: add data/sample composition -- ethinicity, grade, etc -- maybe in table form
%TODO: find out where  the genes were taken from

@@

\subsection{Raw data from Print et al study}
\label{subsec:printrawdat}

%TODO: add data/sample composition -- ethinicity, grade, etc -- maybe in table form
%TODO: find out where  the genes were taken from
%TODO: Change Fuentes-Mattei citation to Cris Print's citation

\subsection{Raw data from \citet{Gatza2010a} study}
\label{sub:gatzarawdat}

%TODO: add data/sample composition -- ethinicity, grade, etc -- maybe in table form
%TODO: find out where  the genes were taken from
%TODO: Change Fuentes-Mattei citation to Cris Print's citation

The raw microarray gene expression data files used in  \citet{Gatza2010a} study were downloaded from the \gls{geo} database (include website name/GEO ID here).
Clinical data for the samples in this study was not available.
130 obesity associated gene probes identified by \citet{Gatza2010a} were downloaded from (supplementary data??).
The raw expression data and the list of gene probes were imported into \gls{r}/RStudio.

@@

\gls{rnaseq} data for the eight cancer types identified in \cref{subsec:tcga} were downloaded from the \gls{icgc} database.
The \gls{rnaseq} data for each of the eight cancer types were imported into \gls{r}/RStudio.

@@

\subsection{Data normalisation}
\label{sub:data_normalisation}

\subsubsection{Microarray data}
\label{ssub:microarray_data}

The raw microarray data were normalised using either the \gls{rma} or \gls{mas} method from the \textit{affy} package in \gls{r}.
Either \gls{rma} or \gls{mas} normalisation was used to account for the inter-array differences that may have arisen during the data collection process.

\subsubsection{RNA-seq data}
\label{ssub:rna_seq_data}

The raw \gls{rnaseq} data for each cancer types downloaded from \gls{icgc} website had 1 added, and then logged to the power of 10.

\subsection{Data standardisation}
\label{sub:data_standardisation}

The data was standardised so that each gene in the data had a mean ($\mu$) of 0 and a standard deviation ($\sigma$) of 1.
This ensured that all of the genes were on the same scale as one another, allowing for comparison across all the genes.

\subsection{Residual data creation}
\label{sub:residual_data_creation}

To create the residual data for a given data set, a linear model was fitted to the data with all of the clinical variables that required to be controlled for.
The \textit{limma} package in \gls{r} was used.

\subsection{Gene probe ID conversion}
\label{sub:gene_probe_id_conversion}

@@

\subsection{Batch correction}
\label{sub:batch_correction}

When there were more than one microarray data to be used as a single data set, these microarray data were normalised separately then combined together with the \texttt{ComBat} function in the \textit{sva} package in \gls{r}.
This corrected for the effect of the individual microarray data on the whole data when they were combined together.

\subsection{Sample \gls{bmi} calculation}
\label{sub:sample_bmi_calculation}

Where missing, the \gls{bmi} of the samples were calculated from the clinical data using \cref{eq:bmicalc}.

\subsection{\gls{icgc} data format conversion}
\label{sub:icgc_data_format_conversion}

The raw \gls{rnaseq} data for each cancer types were formatted using the \textit{dplyr} package in \gls{r}, so that the rows corresponded to the genes and the columns to the samples.

\subsection{Sample ID conversion}
\label{sub:sample_id_conversion}

\gls{icgc} sample ID in the \gls{rnaseq} data were stripped so that the sample ID matched with the \gls{tcga} ID in the clinical data.
After the IDs were stripped, samples were checked for any duplicates.
Where there was a duplicate in the sample ID in the \gls{rnaseq} data, a single sample was chosen to represent that particular ID by using the \texttt{collapseRows} function in the \textit{WGCNA} package in \gls{r}.

\subsection{\gls{icgc} sample exclusion}
\label{sub:icgc_sample_exclusion}

Some of the samples did not have either height or weight data, or both.
These samples were removed from the analyses.
In each of the eight cancer types, the \gls{tcga} sample ID in the clinical data were cross-checked with the sample ID in the \gls{rnaseq} data, and vice versa.
Any sample that did not have either clinical or \gls{rnaseq} data were removed.
This ensured that both clinical and \gls{rnaseq} data were available for all of the samples that were included in the analyses.
See \cref{tab:samplesize} for a summary of the number of samples included in the analyses for each cancer type.

@@

The samples were grouped based on the sample BMI status (obese or non-obese group), and a linear model was fitted to the data based on the sample grouping.
Empirical Bayes statisticsc was used to identify the differentially expressed genes between the two groups, and the top scoring genes were listed.
The error rate of the genes were \gls{fdr} controlled for multiple hypothesis testing.

For a given list of differentially expressed genes identified in a data set, the \texttt{camera} function from the \textit{limma} package was used to identify the pathways that were enriched in those genes.
The \gls{go} database was used to identify the enriched pathways.

\section{\gls{svd}}
\label{sec:singular_value_decomposition}

\gls{svd} was used on a given data set to create the summary metagene scores of a given genetic signature.
% TODO: add more details on the components (s, v, d) created when svd was applied to the data.
Any matrix $X$ can be represented in the form:
\begin{equation}
	\label{eq:svd}
	X = UDV'
\end{equation}

\noindent
where $U$ is the *****, $D$ is the ********, and $V'$ is the **** (eigenvectors and eigenvalues??).

\section{Transformation matrix}
\label{sec:transformation_matrix}

Transformation matrix of a genetic signature was created from the \gls{svd} components.
Rearranging \cref{eq:svd}, you get:

\begin{equation}
	\label{eq:transmat}
	V' = U'D^{-1}X
\end{equation}

\noindent
where $U'D^{-1}$ represents the transformation matrix.
The transformation matrix $U'D^{-1}$ was created from the \gls{svd} components, and was used to transform other data set and obtain the metagene in that data set.

\gls{svd} was applied to the training data set to create the metagene and the transformation matrix in the data set.
The transformation matrix from the training data was used to transform the other data sets and to create the metagene of the genetic signature in those data sets.

Each of these metagenes were plotted in a heatmap to see the association between the metagene and the overall gene expression of the genes that were used to create the metagene.
The metagene was also checked for the association between the sample \gls{bmi} status and \gls{bmi} value, which was plotted in a boxplot and scatter plot, respectively.

\section{Gatza pathway metagene direction}
\label{sec:pathway_metagene_direction}

\section{Obesity metagene prediction}
\label{sec:obesity_metagene_prediction}

@@

% Do I need to explain how to make boxplots and scatter plots?
% \subsection{Box plots}
% \label{sub:box_plots}

% \subsection{Scatter plots}
% \label{sub:scatter_plots}

% TODO: re-write everything so that it's in past tense
